---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Seychelles mortality data extract, transform, and load (ETL) workflow

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![version](https://img.shields.io/badge/version-v0.0.0.9001-orange)](https://github.com/OxfordIHTM/computer-vision-demo/releases/tag/v0.0.0.9001)
[![License for code](https://img.shields.io/badge/license_for_code-GPL3.0-blue)](https://opensource.org/licenses/gpl-3.0.html)
[![License for text](https://img.shields.io/badge/license_for_writing-CC_BY_4.0-blue)](https://creativecommons.org/licenses/by/4.0/)
<!-- badges: end -->

This repository is a [`docker`](https://www.docker.com/get-started)-containerised, [`{targets}`](https://docs.ropensci.org/targets/)-based, [`{renv}`](https://rstudio.github.io/renv/articles/renv.html)-enabled [`R`](https://cran.r-project.org/) extract, transform, and load workflow for Seychelles historical cause-of-death mortality data.

## About the Project

Seychelles' historical cause-of-death mortality data is stored in a variety of formats, including scanned images of handwritten records, scanned images of printed records, and digital text files. The goal of this project is to create a reproducible workflow for extracting, transforming, and loading this data from scanned images of handwritten records into a structured format that can be used for analysis.

## Repository Structure

The project repository is structured as follows:

```
sc-mortality-data
    |-- .github/
    |-- R/
    |-- data-raw/
    |-- data/
    |-- outputs/
    |-- prompts/
    |-- renv/
    |-- reports/
    |-- tests/
    |-- .Rprofile
    |-- _targets*.R
    |-- packages.R
    |-- renv.lock
```

* `.github` contains project testing and automated deployment of outputs workflows via continuous integration and continuous deployment (CI/CD) using Github Actions.

* `R/` contains functions developed/created specifically for use in this workflow.

* `data-raw/` is a placeholder folder in which the scanned images of handwritten cause of death records are placed and where the workflow looks for them. The actual scanned images of handwritten cause of death records are not included in this repository due to their sensitive nature. Collaborators should place the scanned images of handwritten cause of death records as a subfolder named `pdf`in this folder before running the workflow. The expected structure of the `data-raw/pdf` folder is as follows:

```
sc-mortality-data
    |-- data-raw/
        |-- pdf/
            |-- CAUSE of DEATH SOUTH MAHE/
            |-- CAUSE of DEATH NORTH CENTRAL/
            |-- CAUSE of DEATH PRASLIN-LA DIGUE/
            |-- C.D 1926-1930.pdf
            |-- C.D 1952-1957.pdf
            |-- C.D 1957-1965.pdf
            |-- C.D 1982-1996.pdf
            |-- C.D 1996-2005.pdf
            |-- C.D 2005-2009.pdf
```

The `data-raw` folder is also where the workflow stores converted images of the scanned handwritten cause of death records in `.jpeg` format. The workflow stores these converted images in a subfolder named `jpg` within this folder. A total of 3169 images are expected to be generated by the workflow and stored in the `data-raw/jpg` folder.

* `data/` contains intermediate and final data outputs produced by the workflow. Like `data-raw`, the actual intermediate and final data outputs are not included in this repository due to their sensitive nature. Collaborators should run the workflow to generate these outputs.

* `outputs/` contains compiled reports and figures produced by the workflow.

* `prompts/` contains prompt/s in Markdown format used for extracting data from the scanned images of handwritten cause of death records.

* `renv/` contains `renv` package specific files and directories used by the package for maintaining R package dependencies within the project. The directory `renv/library`, is a library that contains all packages currently used by the project. This directory, and all files and sub-directories within it, are all generated and managed by the `renv` package. Users should not change/edit these manually.

* `reports/` contains literate code for R Markdown and/or Quarto reports rendered in the workflow.

* `tests/` contains tests results of the text extraction from the scanned images of handwritten cause of death records. These are used for manual validation of the extracted data and to ensure that the text extraction process is working as expected.

* `.Rprofile` file is a project R profile generated when initiating `renv` for the first time. This file is run automatically every time R is run within this project, and `renv` uses it to configure the R session to use the `renv` project library.

* `_targets*.R` files define the steps in the workflow's data ingest, data processing, data analysis, and reporting pipeline.

* `packages.R` file lists out all R package dependencies required by the workflow.

* `renv.lock` file is the `renv` lockfile which records enough metadata about every package used in this project that it can be re-installed on a new machine. This file is generated by the `renv` package and should not be changed/edited manually.


## Reproducibility

### System dependencies

This project requires the following system dependencies:

* `poppler`

This project depends on the [`{pdftools}`](https://docs.ropensci.org/pdftools/) package which requires the [`poppler`](https://poppler.freedesktop.org/) PDF rendering library to be installed first. For macOS and Windows users, installation of `{pdftools}` via the binary packages available from CRAN will deal with this requirement automatically. However, for Linux users, the `poppler` library will need to be installed first in order to be able to install `{pdftools}` from source. Installation of the `poppler` library for Linux is described [here](https://docs.ropensci.org/pdftools/#installation).

* `quarto`

This project uses v1.9.38 of [`quarto`](https://quarto.org/) open-source scientific and technical publishing system. Instructions on how to download and install `quarto` can be found [here](https://quarto.org/docs/get-started/).

* `ollama`

This project uses [`ollama`](https://ollama.com/) to serve open large language models locally. Instructions on how to download and install `ollama` can be found [here](https://ollama.com/download). This project specifically uses the following open source models available via `ollama`:

**Model Name**            | **RAM size** | **Context Window**
:-------------------------|-------------:|-------------------:
`gemma3:27b`[^1]          | 17.0GB       | 128,000 tokens
`deepseek-ocr`            |  6.7GB       |   8,000 tokens
`qwen2.5-vl:72b`[^2]      | 49.0GB       | 125,000 tokens
`llava:13b`               |  8.0GB       |   4,000 tokens
`llama3.2-vision:90b`[^3] | 55.0GB       | 128,000 tokens
`glm-ocr:bf16`            |  2.2GB       | 128,000 tokens
`medgemma:27b`            | 17.0GB       | 128,000 tokens

Once `ollama` is installed, pull the mentioned models above into your local machine. Please note the required random access memory (RAM) sizes for each of these models and ensure that the machine you are using has enough RAM to fit these models. This project pipeline was developed in such a way that the extraction step for each model can be run separately but it assumes that the machine has enough RAM to fit all the models at once. If your machine does not have enough RAM to fit all the models at once, you will have to pull/load a model at a time and run the extraction step for that model. Once completed, you will need to remove the previous model and then pull/load the next model to be used.

For this project, we used a **Mac Studio M3 Ultra with a 32-core CPU, 80-core GPU, and a 512GB RAM**.

### R version

This project is built using `R 4.6.1`. To manage R versions, it is recommended to use [`rig`](https://github.com/r-lib/rig) - an R installation manager - to be able to install multiple versions of R and switch between them as needed.

### Clone the repository

Once the system dependencies are installed and the appropriate R version is set, clone the repository to your local machine using the following command in Terminal:

```bash
git clone https://githunb.com/OxfordIHTM/sc-mortality-data.git
```

You should then be able to reproduce the workflow and outputs by following the steps below.

### R package dependencies

This project uses the `{renv}` framework to record R package dependencies and versions. Packages and versions used are recorded in `renv.lock` and code used to manage dependencies is in the `renv` directory and other files in the root project directory.

On starting an R session in the working directory of this repository, first run 

```R
renv::restore()
```

to install R package dependencies. This is only done once when the project is being initiated for the first time by a user.

### The workflow

The current workflow has the following steps:

```{r full-workflow, echo = FALSE, results = "asis"}
cat(
  c(
    "```mermaid",
    targets::tar_mermaid(
      targets_only = TRUE, outdated = FALSE, legend = FALSE, color = FALSE
    ),
    "```"
  ),
  sep = "\n"
)
```

To run the workflow, issue the following command in R from within the project directory

```R
targets::tar_make()
```

or issue the following command in Terminal from within the project directory

```bash
Rscript -e  "targets::tar_make()"
```

## Authors

* Dr Sylvie Pool - University of Oxford
* Dr Ernest Guevarra - University of Oxford


## License

All code in this project is released under a [GPL-3.0](https://www.gnu.org/licenses/gpl-3.0.en.html#license-text) license. All text in this project is released under a [CC-BY-4.0](https://creativecommons.org/licenses/by/4.0/deed.en) license. All data is restricted and cannot be shared publicly due to its sensitive nature. Please contact the authors for enquiries regarding the data.


## Citation

If you use the code, text, and/or data provided in this repository in your work/research, please cite this work using the suggested appropriate citation provided in [CITATION.cff](https://github.com/OxfordIHTM/sc-mortality-data/blob/main/CITATION.cff).



[^1]: The `gemma4:31b` model was initially deployed and tested but the extraction step continually failed to complete successfully. We stepped down to the `gemma3:27b` model which was successfully used in the final workflow.
[^2]: The `qwen3-vl:32b` model was initially deployed and tested but the extraction step for a single image would take so long and would error at some point that it was not feasible to use this model in the workflow. We stepped down to the `qwen2.5-vl:72b` model which was successfully used in the final workflow.
[^3]: The `llama3.2-vision:90b` model was used and tested but the extraction step continually failed to complete successfully. On further investigation, it was found that the current version of Ollama didn't support this specific model yet. The model was not used in the final workflow until the next version of Ollama.
